\normalsize{Biometry of the invasive lionfish (\textit{Pterois volitans}) in

### 2017-07-29 Manuscript.Rmd
Added study area section and some discussion

---
title: \normalsize{Biometry of the invasive lionfish (\textit{Pterois volitans}) in
  Playa del Carmen, Mexico and a review of length - weight parameters across the invasion
  range}
csl: apa.csl
fontsize: 12pt
output:
  pdf_document:
    fig_caption: yes
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
bibliography:
- references.bib
- packages.bib
---

@@

At least 84% of the marine eco-regions have reported the presence of an invasive species [@molnar_2008]. The presence of invasive alien species represents a major threat to local biodiversity and the economic activities that depend on it, like tourism or fisheries [@bax_2003]. By 2005, the economic cost of invasive species to the United States was  estimated at \$120 billion per year and nearly 42% of species that have been included in the Endangered or Threatened species list have been labeled as such due to presence of invasive species [@pimentel_2005]. Invasive species may threaten native species through competition [@davis_2003] or predation. This highlightes the importance of understanding, managing, and preventing ecological invasions.

Lionfish (*Pterois volitans - miles* complex) are an invasive species in the North-Western Atlantic and the Caribbean likely introduced through libreation of aquarium-kept organisms [@betancurr_2011]. They are the first marine vertebrates to establish in North Atlantic [@schofield_2009: @schofield_2010] and Caribbean coasts. Lionfish have been recorded in coral reefs [@aguilarperera_2010], but also in other habitats like estuaries [@jud_2011], mangroves [@barbour_2010] and hard-bottom reefs [@muoz_2011]. Their presence in these waters has been labeled as a major marine invasion due to its threat to local biodiversity, the veolocity of its invasion, and its difficulty of management [@hixon_2016].

As ways to reduce the abundance of this invasive species local authorities have promoted removal programs incentivized consumption of this invasive species [@chin_2016]. Lionfish removals have shown to significantly reduce -but not quite eliminate- lionfish abundances at local scales [@sandel_2015, @chin_2016, @deleon_2013]. While complete erradication through fishing effort is unlikely to be a solution due to the rapid recovery rates exhibited by the species [@barbour_2011], further incentivizing its consumption might create a market demand big enough to promote a fishery [@chin_2016].

A significant amount of research has been done on lionfish feeding ecology in North Carolina [@muoz_2011], the Bahamas [@morris_2009; @cote_2013], Northern Gulf of Mexico [@dahl_2014], Mexican Caribbean [@valdezmoreno_2012; @villaseorderbez_2014], and Costa Rica [@sandel_2015]. It has been identified that lionfish can reduce recruitment [@albins_2008] and population sizes [@green_2012] of native reef-fish species, and further the endangerement of critically endangered reef fish [@rocha_2015]. This information has been useful in predicting the trophic impacts of lionfish [@ariasgonzalez_2011]. Significant efforts have also been made to understand the possible impacts of the invasion by keeping track of the invasion range through time [@schofield_2009; @schofield_2010] and predicting invasion ranges under climate change scenarios (@grieve_2016).

Biomass is a standard indicator used to measure the state or size of a population. Biomass has been used in lionfish research to convert individual fish length observations into fish weight. In order to do this, researchers must find growth parameters thad describe the growth of their species for a particular region or location. Having precision of these parameters is important, because small changes in any of them can result in important over- or underestimation of biomass levels. This becomes particularly important when research involves, for example, identifying the total biomass available for harvest by fishers -as invasion management programs- or the efficacy of lionfish removals.

The objectives of this study include providing site-specific weight-at-length parameters for the invasive lionfish in the Mexican Caribbean region. At the same time, we highlight the importance of using site-specific parameters by estimating biomass with parameters from other regions across the invasion range and comparing them to observed biomasses. Finally, we summarize other weight-at-length studies presented, making them readily accessible for future research.

@@

The present study took place off the coasts of Playa del Carmen, in the coasts of the Mexican Caribbean. The region represents the northernmost section of the Mesoamerican reef, which extends from coast of Cancun south to Honduras (**reference**). Coral reefs and mangroves are important habitats distributed along the cost, and represent important sources of income in terms of extractive (*e.g.* recreational fishing) and non-extractive (*e.g.* SCUBA diving) activities related to tourism (**reference**), the main source of income to the local economy.

Coral reefs in the region are characterized by X, Y, Z. Descripcion general de los arrecifes.

@@

pen_map <- tm_shape(stat, bbox = bb(xlim = c(-89.29656, -86.71074), ylim = c(17.89399, 21.60525))) +
  tm_polygons() +
  tm_shape(res1) +
  tm_symbols(shape = 5, size = 2, border.col = "black", col = "black", border.lwd = 2) +
  tm_grid() +
  tm_scale_bar()

@@

The present study used samples obtained by @villaseorderbez_2014 for stomach content analysis between May and August of 2010. A total of 33 SCUBA immersions were performed in `r length(unique(fish$Location))` sampling sites along 14 Km of coast between Puerto Aventuras and Akumal (Fig. 1, Table 1). Sampling locations included wall and carpet reefs at depths between `r min(fish$Depth_m)` m and `r max(fish$Depth_m)` m.  All observed organisms (n = `r dim(fish)[1]`) were collected using hand nets -to avoid weight loss due to bleeding- and numbered collection bottles. Information on depth and other comments were recorded in an underwater slide. Depth was recorded by dive gauges held by divers as safety procedures during the collections. Samples were frozen within 30 minutes of completing the dive and stored for posterior analysis in the lab. For every organism, Total Length (TL; mm) and Total Weight (TW; gr) were recorded in the lab.

@@

All hypothesis testing was performed with a confidence level of $\alpha = 0.01$ and performed in R version 3.4.0 [@rcore_2017]. Data wrangling was performed with the tidyverse package [@tidyverse_2017]. Maps were created with a mix of functions from the sp [@sp_2017], rgdal [@rgdal_2017], tmap [@tmap_2017], and tmaptools [@tmaptools_2017] packages. Correction for heteroskedastic-robust standard errors was done with the sandwich [@sandwich_2014] and lmtest [@lmtest_2002] packages. Models were manipulated with with the broom package [@broom_2017]. Stargazer [@stargazer_2015] was used to produce the tables, and refmanager [@RefManager_2014] was used to keep track of citations. The manuscript was written in rmarkdown [@rmarkdown_2017] and processed with the knitr package [@knitr_2017].

@@

@@

>>>>

---
title: \normalsize{Biometry of the invasive lionfish (\textit{Pterois volitans}) in
  Playa del Carmen, Mexico and a review of allometric growth parameters across the invasion
  range}
fontsize: 12pt
output:
  pdf_document:
    fig_caption: yes
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
bibliography:
- references.bib
- packages.bib
---

@@

At least 84% of the marine eco-regions have reported the presence of an invasive species [@molnar_2008]. These represents a major threat to local biodiversity and the economic activities that depend on it, like tourism or fisheries [@bax_2003]. Invasive species may also threaten native species through competition [@davis_2003] or predation. By 2005, the economic cost of invasive species to the United States was  estimated at \$120 billion per year and nearly 42% of species that have been included in the Endangered or Threatened species list have been labeled as such due to presence of invasive species [@pimentel_2005]. This highlightes the importance of understanding, managing, and preventing ecological invasions.

Lionfish (*Pterois volitans - miles* complex) are an invasive species in the North-Western Atlantic and the Caribbean, likely introduced through libreation of aquarium-kept organisms [@betancurr_2011]. They are the first marine vertebrates to establish in North Atlantic [@schofield_2009; @schofield_2010] and Caribbean coasts [@sabidoitza_2016]. Lionfish have been widely reported in coral reefs [@aguilarperera_2010], but also in other habitats such as estuaries [@jud_2011], mangroves [@barbour_2010], areas with hard-bottoms [@muoz_2011], and mesophotic reefs [@andradibrown_2017]. Due to its threat to local biodiversity, the speed of their spread, and its difficulty of management, their presence in these waters has been labeled as a major marine invasion  [@hixon_2016].

A significant amount of research has been done to describe lionfish feeding ecology in North Carolina [@muoz_2011], the Bahamas [@morris_2009; @cote_2013], Northern Gulf of Mexico [@dahl_2014], Mexican Caribbean [@valdezmoreno_2012; @villaseorderbez_2014], and Costa Rica [@sandel_2015]. Their feeding behaviour and high consumption rates can reduce recruitment [@albins_2008] and population sizes [@green_2012] of native reef-fish species, and further the endangerement of critically endangered reef fish [@rocha_2015]. Major efforts have also been made to understand the possible impacts of the invasion by keeping track of its range through time [@schofield_2009; @schofield_2010] and predicting invasion ranges under climate change scenarios (@grieve_2016). By combining these information, researchers have been ableto predict the trophic impacts of lionfish [@ariasgonzalez_2011], which can then be translated into ecosystem-level impacts.

Seeking to reduce the abundance of this invasive species, local authorities have promoted removal programs ands incentivized its consumption [@chin_2016]. Lionfish removals have shown to significantly reduce -but not quite eliminate- lionfish abundances at local scales [@sandel_2015, @chin_2016, @deleon_2013]. The rapid recovery rates exhibited by lionfish make of complete erradication through fishing effort an unlikely solution [@barbour_2011]. However, further incentivizing its consumption might create a demand big enough to promote and sustain a stable fishery [@chin_2016], which can reduce local abundances and control -not erradicate- the invasion.

Research focused on erradication and management effectivenes must be based on sound biological information, highlighting the importance of both fields. Management effectiveness often identified the gains associated to the invested effort, like biomass extracted per dollar. Among fish, biomass is often used as a standard measure of the state or size of a population. By deffinition, biomass is the sum of the weight of all organisms. In some cases, a fish's weight is directly measured, in others,  weight can be calculated from length data. In order to do this, researchers must find allometric parameters that describe the weight-at-length for their species of interest at a particular region. Having precision of these parameters is important, because small changes in any of them can result in over- or underestimation of an organisms weight, and therefore total biomass. This becomes particularly important when research involves, for example, identifying the total biomass available for harvest by fishers [@chin_2016] or the efficacy of lionfish removals [@barbour_2011].

Here, I provide site-specific allometric growth parameters for the invasive lionfish in the central Mexican Caribbean region, where no studies have been published. At the same time, I highlight the importance of using site-specific parameters by estimating biomass with parameters from other regions across the invasion range and comparing them to observed biomass. Finally, I provide a review of other weight-at-length studies -making them readily accessible for future research- and provide a set of guidelines to standardize the way in which allometric parameters should be reported to facilitate their use.

@@

The study took place off the coasts of Playa del Carmen, in the Mexican Caribbean (**Fig. 1**). The region represents the northernmost section of the Mesoamerican Barrier Reef System [@ruiz-zarate_2003]. Coral reefs and mangroves are localy important habitats that represent important sources of income in terms of extractive (*e.g.* recreational fishing) and non-extractive (*e.g.* SCUBA diving) activities related to tourism, the main source of income to the local economy [@murray_2007].

In the region, reef profile has been described by @ariasgonzales_1998, where the reef lagoon extends about 500 m from the coast, until the reef crest is reached. The reef becomes deeper, leading to the reef front often found at 700 m from the coastline end extends for an additional 300 m. At approximately 1000 m away from shore and 30 - 40 m depth, the reef leads to a drop-off. Along a perpendicular profile to the coast, bands of reef are interrupted by sand patches at 8 - 12 m deep and 16-18 m deep. Along the coast, these reefs have been reported to be under significant antrhopogenic pressure, likely causing a shift in structure and function [@Bozec_2008].

@@

pen_map <- tm_shape(stat, bbox = bb(xlim = c(-89.29656, -86.71074), ylim = c(17.89399, 21.60525))) +
  tm_polygons() +
  tm_shape(res1) +
  tm_symbols(shape = 5, size = 2, border.col = "black", col = "black", border.lwd = 2) +
  tm_scale_bar()

@@

Organisms sampled by @villaseorderbez_2014 for stomach content analysis between May and August of 2010 were used in this work. A total of 33 SCUBA immersions were performed in `r length(unique(fish$Location))` sampling sites along 14 Km of coast between Puerto Aventuras and Akumal (Fig. 1, Table 1). Sampling locations included wall and carpet reefs at depths between `r min(fish$Depth_m)` m and `r max(fish$Depth_m)` m.  All observed organisms (n = `r dim(fish)[1]`) were collected using hand nets -to avoid weight loss due to bleeding- and numbered collection bottles. Information on depth and other comments were recorded in an underwater slide. Depth was recorded by dive gauges held by divers as safety procedures during the collections. Samples were frozen within 30 minutes of completing the dive and stored for posterior analysis in the lab. For every organism, Total Length (TL; mm) and Total Weight (TW; gr) were recorded in the lab.

@@

All hypothesis testing was performed with a confidence level of $\alpha = 0.01$ in R version 3.4.0 [@rcore_2017]. Data wrangling was performed with the tidyverse package [@tidyverse_2017]. Maps were created with a mix of functions from the sp [@sp_2017], rgdal [@rgdal_2017], tmap [@tmap_2017], and tmaptools [@tmaptools_2017] packages. Callculation of heteroskedastic-robust standard errors was done with the sandwich [@sandwich_2014] and lmtest [@lmtest_2002] packages. Models were manipulated with with the broom package [@broom_2017]. Stargazer [@stargazer_2015] was used to produce the tables, and refmanager [@RefManager_2014] was used to keep track of citations. The manuscript was written in rmarkdown [@rmarkdown_2017] and processed with the knitr package [@knitr_2017]. Raw data and code used in this work is available at [github.com/jcvdav/lionfish_biometry](github.com/jcvdav/lionfish_biometry).

@@

Here, I provide site-specific allometric growth parameters for the invasive lionfish in the central Mexican Caribbean region, where no studies have been published.

Other studies have provided allometric growth parameters of lionfish in Mexican waters, specifically for the Alacranes Reef, in the Gulf of Mexico [@aguilarperera_2018], and further south in the Caribbean for the Xcalak National Park [@sabidoitza_2016], but no information for the central region of the Mexican Caribbean coast had been provided. Furthermore by using hand nets instead of spears, we are able to sample a wider range of sizes, extending the lower length classes to smaller organisms often not sampled with spears. Thus, this study is the first to provide allometric growth parameters for lionfish in the central region of the Mexican Caribbean and includes a wider range of total lengths, allowing us to calculate -with certainty- a wider range of weights.

The parameters estimated in this study were within the range of studies in other regions, however using other parameters resulted in important differences in expected biomass. These differences can have major implications in management, especially when estimating biomass availabe for harvest or predicting population size and effects on local ecosystems. Using site-specific values provides a more aqurate estimate of fish biomass. Future research should try to use, to the extent possible, parameters calculated for their region, or use different parameters to provide upper and lower bounds in their results. At the same time, this higlights the need for mor basic research that can describe allometric growth parameters -and other biologically important information- of lionfish throughout its invasion range.

While performing the lietrature review, it was often unclear if parameters presented for **eq.1** or **eq. 3**. Sometimes, they were even misslabeled and yielded senseless results when using the suggested conversion equation. On some others, parameters were said to be reported for mm to gr conversions, when they were actually reported as cm to gr conversions. Perhaps these minor discrepancies can be easily solved by any researcher, but why should they exist in the first place? It is important that we report our information in a standard way, making it readily available for other researchers but also managers. In this particular case, I provide my humble opinion in X simple guidelines to report allometric parameters:

1.

2. Be explicit in the methods section. What may seem obvious to you as an author -because you have been deeply immersed throughout the process- may not be clear to the reader. Specify any transformation performed on the data; When using log-transformations, mention the base used to transform. Do not asume that "data were log-transformed" means $log_{10}(X)$. These assumptions vary across disciplines and softwares. For example, in biology we often assume log-transformed indicates the use of base 10, however in R the command is log10() and not log(), which uses base $e$.

3. Use heteroskedastic-robust standard errors.

4. At the least, specify the equation for which parameters are presented and try include an explicit example with the parameters substituted into it, as done by [X Y y Z]. If possible, present the parameters in their exponential and linear form.

5. Report standard errors and confidence intervals arround the obtained estimates. Given that small changes in $a$, $c$,  and $b$ can result in important changes in estimated weight, it is important that we report uncertainty arround each parameter and not just general model fit (*i.e.* R^2^). Additionally, make every possible effort to report at least 4 decimal points. If the journal has style guidelines about this, include full estimates in a table as supplementary material.

6. Make your data -and code- available. Even if this is not requited by the journal or publisher, you can use free cloud data storage services to make your research accessible to others. In times and countries where resources are limited, it is important that we take advantage of all open science tools that promote the advancement of knowledge. Ultimately, this fosters transparency, allows replicability of research, and can lead to collaborations.

While in this work I make reference to allometric growth parameters, these guidelines can easily be applied to other areas and disciplines. It is important

@@

Finally, I provide a review of other weight-at-length studies -making them readily accessible for future research- and provide a set of guidelines to standardize the way in which allometric parameters should be reported to facilitate their use.
